---
title: "Final Paper Citations"
author: "Micah Clark Moody"
date: '2022-07-29'
output: html_document
---

### FLAG: Add a TOC

# Final Paper Citations
This `.Rmd` document has the code for quantitative claims in my final draft of the MA Thesis *“Bond is going to be…”: Setting and justifying pretrial conditions in Cook County Central Bond Court*. For each claim, I'll include the quote and code.

## DONE Setup r environment
Import needed packages to run code in the r environment.
```{r setup, message=FALSE}
library(knitr)
library(lubridate)
library(readr)
library(tidyverse)
```

## DONE Import data
Import data needed to run code in the r environment. This analysis only relies on the data I recorded during my obsevations, available in raw form in the "data" folder of this github repository linked [here](https://github.com/MicahCM/bond_court/blob/main/data/data_cleaned.csv). The raw file is available [here](https://raw.githubusercontent.com/MicahCM/bond_court/main/data/data_cleaned.csv) and directly inked in the code.

```{r, message=FALSE}
# importing data from my observations
# url link is from this github repository
obs_data <- read.csv(url("https://raw.githubusercontent.com/MicahCM/bond_court/main/data/data_cleaned.csv"))

# coercing numeric variables
obs_data$can_post <- as.numeric(obs_data$can_post) 
obs_data$bond_amount <- as.numeric(obs_data$bond_amount)

# creating  variable for unaffordable bond
# unaffordable bond is defined as can_post is less than bond_amount
d_bond_detail <- obs_data %>%
  filter(bond_type == "D") %>%
  mutate("bond_type_detail" = ifelse(((can_post * 10) < bond_amount), 
                                     "D_affordable", "D_unaffordable"))

# creating new variable "bond_type_detail" that reflects D bond affordability
obs_data <- obs_data %>%
  filter(bond_type != "D") %>%
  mutate("bond_type_detail" = bond_type) %>%
  rbind(d_bond_detail)

# factoring variables to organize visualization
# creating levels
class_order <- c("4", "3", "2", "1", "X")
bond_order <- c("I", "D", "No bond")
bond_detail_order <- c("I", "D_affordable", "D_unaffordable", "No bond")

# mutating columns
obs_data <- obs_data %>%
  mutate(class_1 = factor(class_1, 
                          levels = class_order)) %>%
  mutate(bond_type = factor(bond_type,
                            levels = bond_order)) %>%
  mutate(bond_type_detail = factor(bond_type_detail,
                            levels = bond_detail_order))
  
```

## DONE Dismissing low-level drug cases
__Page 15: "In bond court, ASAs dismiss almost one in five low-level drug cases."__

- "Low-level" is represented by `class_1 == 4` meaning it is a class 4 felony charge
- "drug case" is represented by `charge_type` which I created as a manual grouping variable
- "dismiss" is represented by `nole`, a dummy variable for when cases are dismissed. 

Code interpretation: 16% of class 4 drug cases are dismissed by prosecutors

```{r dismissing low-level drug cases, message=FALSE}
obs_data %>%
  
  # selecting low-level drug cases
  filter(
    class_1 == 4 & (
      charge_type == "4 - drug sales (manufacture/delivery, PCS with intent)" | 
      charge_type == "3 - drug possession (PCS)")) %>%
  
  # counting dismissed cases (nole)
  group_by(nole) %>%
  summarize(n()) %>%
  
  # calculating percent dismissed and prosecuted
  mutate("percent" = ((`n()`/(50))*100)) %>%
  
  # printing results
  kable(col.names = c("Dismissed Case", "Total Count", "Percentage"), caption = "Low-Level (class 4) Drug Case Dismissal Rates")
```

## DONE Release in low-level cases
__Page 19: "Affordable bail, money bail less than or equal to the stated ability to pay, is most common in low-level (class 4) felony cases."__

__Page 19: "In low-level cases I observed, the pretrial condition was release without money bail in just under half of cases."__

```{r release in cases, message=FLASE}
# summing each bond type, separated by D bond affordability
bond_summary <- obs_data %>%
  filter(bond_type_detail != ""
         & class_1 %in% c("1", "2", "3", "4", "X")) %>%
  group_by(class_1, bond_type_detail) %>%
  summarize(total = n())

# summing by class 
bond_class_summary <- aggregate(bond_summary$total,
            by = list(class_1 = bond_summary$class_1),
            FUN = sum)

# adding factors to data frame
bond_summary %>% 
  left_join(bond_class_summary) %>%
  mutate(percentage = total / x) %>%
  
  # visualizing results
  ggplot(mapping = aes(
    x = class_1,
    y = percentage,
    fill = bond_type_detail
  )) + 
    geom_col(
      position = position_dodge()
    )
```

## Gun possession with previous conviction(s)
__Page 20-21: "For example, release on money bail and some type of supervision is the common pretrial condition for defendants charged with gun possession who have a previous gun conviction in their history."__

```{r}
obs_data %>%
  filter(charge_type == "6 - gun possession (prior record)") %>%
  ggplot(mapping = aes(
    x = bond_type,
    fill = condition
  )) +
  geom_bar(position = position_dodge())
```

## DONE Class X gun possession pretrial conditions
__Page 24-25: Of the class X gun possession cases I observed, prosecutors requested pretrial jailing half of the time.__
```{r}
obs_data %>%
  
  # selecting class X gun possession case
  # called "armed habitual criminal" which I abbreviated as "AHC" in the charge variable
  filter(charge_1 == "AHC") %>%
  
  # graphing results
  ggplot(mapping = aes(
    x = bond_type
  )) + 
  geom_bar()
```

## DONE FOID card possession and release conditions
__Page 28: In a third of class four gun arrests, the defendant is a legal gun owner with a FOID card Illinois gun license__

```{r}
obs_data %>%
  
  # isolating class 4 gun arrests where the gun license status is recorded
  # in 19 class 4 gun cases I did not record the gin license status
  filter(class_1 == 4, charge_1 == "UUW", gun_license != "") %>%
  
  # graphing data
  ggplot(mapping = aes(
    x = gun_license
    )) +
  geom_bar()
```

__Page 28: In the cases I observed, nine of ten defendants with a FOID card are released on an I-Bond, the most lenient pretrial condition. In contrast, only two of ten defendants without a FOID card and arrested on gun charges were released on an I-Bond with most given money bail as a pretrial condition.__

```{r}
obs_data %>%
  
  # isolating class 4 gun arrests where the gun license status is recorded
  # in 19 class 4 gun cases with license status I did not record bond type in 4 cases
  filter(class_1 == 4, 
         charge_1 == "UUW", 
         gun_license %in% c("FOID no CCL", "No FOID or CCL"), 
         bond_type != "") %>%
  
  # graphing data
  ggplot(mapping = aes(
    x = bond_type
  )) +
  geom_bar() +
  facet_wrap(~gun_license)
```

## DONE Drug charges pretrial conditions set overview

__Page 32: Judges release two-thirds of defendants with drug charges...__ 

- Release means 1) an I-Bond was set or 2) an affordable D-Bond was set

__Page 32: I observed consistent money bail and pretrial supervision conditions between both charge categories__

```{r}
obs_data %>%
  filter(charge_type %in% c("3 - drug possession (PCS)", 
                            "4 - drug sales (manufacture/delivery, PCS with intent)")
         
         # I did not record bond type for 27 drug case observations 
         & bond_type_detail != "")  %>%
  mutate(bond_type_detail = factor(bond_type_detail,
                            levels = bond_detail_order)) %>%

  # data visualization
  ggplot(mapping = aes(
    x = bond_type_detail
  )) + 
  geom_bar() + 
  facet_wrap(~ charge_type)
```

