---
title: "Final Paper Citations"
author: "Micah Clark Moody"
date: '2022-07-29'
output: 
  github_document:
    toc: true
---
This `.Rmd` document has the code for quantitative claims in my final draft of the MA Thesis *“Bond is going to be…”: Setting and justifying pretrial conditions in Cook County Central Bond Court*. For each claim, I'll include the quote and code.

# Setup r environment
Run this section to import packages into the r environment. For this section, and all sections, the `.Rmd` file for code.
```{r setup, warning = FALSE, message = FALSE, echo = FALSE}
# muting warnings and messages in output file globally
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)

# importing needed packages
library(knitr)
library(readr)
library(tidyverse)
```

# Import data
In this section, I import and clean the data from my observations. This data is available in the "data" folder of this github repository linked [here](https://github.com/MicahCM/bond_court/blob/main/data/data_cleaned.csv). The raw file is available [here](https://raw.githubusercontent.com/MicahCM/bond_court/main/data/data_cleaned.csv) and directly inked in the code.

```{r import data}
### IMPORTING DATA
# url link is from this github repository
original_data <- read.csv(url("https://raw.githubusercontent.com/MicahCM/bond_court/main/data/data_cleaned.csv"))

# creating a duplicate dataset for clearning
obs_data <- original_data

### SETTING VARIABLE FOR AFFORDABLE AND UNAFFORDABLE MONEY BAIL
# coercing numeric variables needed to calculate bail affordability
obs_data$can_post <- as.numeric(obs_data$can_post) 
obs_data$bond_amount <- as.numeric(obs_data$bond_amount)

# creating a variable for unaffordable bond
# unaffordable bond is defined as can_post is less than bond_amount
d_bond_detail <- obs_data %>%
  filter(bond_type == "D") %>%
  mutate("bond_type_detail" = ifelse(((can_post * 10) < bond_amount),
                                     "D_affordable", "D_unaffordable"))

# creating new variable "bond_type_detail" that reflects D bond affordability
obs_data <- obs_data %>%
  filter(bond_type != "D") %>%
  mutate("bond_type_detail" = bond_type) %>%
  rbind(d_bond_detail)

### FACTORING VARIABLES TO ORGANIZE DATA VISUALIZATIONS
# creating levels
class_order <- c("4", "3", "2", "1", "X")
bond_order <- c("I", "D", "No bond")
bond_detail_order <- c("I", "D_affordable", "D_unaffordable", "No bond")

# applying levels to variables in the data
obs_data <- obs_data %>%
  mutate(class_1 = factor(class_1, 
                          levels = class_order)) %>%
  mutate(bond_type = factor(bond_type,
                            levels = bond_order)) %>%
  mutate(bond_type_detail = factor(bond_type_detail,
                            levels = bond_detail_order))

# saving color scale based on levels
bond_detail_color <- c("deepskyblue", "grey60", "grey25", "lightcoral")
```

# Dismissing low-level drug cases
> _"In bond court, ASAs dismiss almost one in five low-level drug cases"_ (Page 15)

As the data visualization shows, 16% of class 4 drug cases are dismissed by prosecutors. In this analysis, "low-level" is defined as a class 4 felony charge. In the data, the variable `nole` is a dummy for case dismissal where 1 = dismissed and 0 = no dismissed. The name `nole` is an abbreviation of the phrase "nolle prosequi," a Latin legal term that prosecutors say when dismissing a case.

These results are likely an under count of dismissed cases because judges do not always announce the charges before prosecutors move to dismiss the case, meaning I observed cases that were dismissed without identifying the underlying charge. 

```{r dismissing low-level drug cases, message = FALSE}
### FINDING TOTAL CLASS 4 DRUG CASES
# using observation data
class_4_drug_case_total <- obs_data %>%
  
  # selecting low-level (class 4) drug cases
  filter(
    class_1 == 4 & (
      charge_type == "4 - drug sales (manufacture/delivery, PCS with intent)" | 
      charge_type == "3 - drug possession (PCS)")) %>%
  
  # counting total cases
  count()

### CALCULATING DISMISSAL PERCENTAGE
# using observation data
obs_data %>%
  
  # selecting low-level (class 4) drug cases
  filter(
    class_1 == 4 & (
      charge_type == "4 - drug sales (manufacture/delivery, PCS with intent)" | 
      charge_type == "3 - drug possession (PCS)")) %>%
  
  # counting dismissed cases (nole)
  group_by(nole) %>%
  summarize(n()) %>%
  
  # calculating percent dismissed and prosecuted
  mutate("percent" = `n()`/(class_4_drug_case_total$n)) %>%
  mutate(nole = ifelse(nole == 1, "Dismissed", "Not Dismissed")) %>%
  
  # visualizing results
  ggplot(mapping = aes(
    x = nole,
    y = percent
  )) + 
  
  # controlling graph aesthetics
  geom_col(fill = "grey25") +
  theme_minimal() + 
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Figure 1: Percentage of Drug Cases Dismissed During Central Bond Court",
    caption = "Central Bond Court Observations, January 2022 to May 2022") +
  xlab("") + 
  ylab("")
```

#  Release in low-level cases 
> _"Affordable bail, money bail less than or equal to the stated ability to pay, is most common in low-level (class 4) felony cases"_ (Page 19)

From Figure 2 we can see that over 52% of class 4 cases lead to release on an I-Bond. Further, when and I-Bond is not set, an affordable money bail is the next most common pretrial condition. For class 4 cases, release through an I-Bond or affordable money bail is most common, occuring in 76% of cases I observed. 

Class 3 cases also frequently lead to release on an I-Bond or affordable money bail. However, we can see in Table 1 the level of unaffordable money bail is slightly higher, making release slightly less common in class three cases. 

Class 2, 1, and X charges all have lower rates of release.

This sentence could have been more precisely phrased "[r]elease through no bail or affordable bail, money bail less than or equal to the stated ability to pay, is most common in low-level (class 4) felony cases." This would be more precise because class 3 cases have a higher rate of affordable bail with 58% of cases. there is a lower rate of pretrial release overall, however, at 70% compared to the 76% release rate for class 4 felony cases.

```{r release in low-level cases}
# counting case by class and detailed bond type
bond_summary <- obs_data %>%
  filter(bond_type_detail != ""
         & class_1 %in% c("1", "2", "3", "4", "X")) %>%
  group_by(class_1, bond_type_detail) %>%
  summarize(total = n())

# counting case by class only
bond_class_summary <- aggregate(bond_summary$total,
            by = list(class_1 = bond_summary$class_1),
            FUN = sum)

# calculating percentage of bond type grouped by class
bond_summary_full <- bond_summary %>% 
  left_join(bond_class_summary) %>%
  mutate(percentage = total / x)

### CALCULATING THE OBSERVATIONS LOST BY DEFINING BOND AFFORDABILITY
d_bond_unfiltered <- original_data %>%
  filter(bond_type == "D") %>%
  count()

d_bond_filtered <- obs_data %>%
  filter(bond_type_detail %in% c("D_affordable", "D_unaffordable")) %>%
  count()

### VISUALIZING PERCENTAGES BY CLASS AND BOND TYPE
bond_summary_full %>%
  
  # creating visualization
  ggplot(mapping = aes(
    x = class_1,
    y = percentage,
    fill = bond_type_detail
  )) + 
  
  # controlling graph aesthetics
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) + 
  scale_fill_manual(values = bond_detail_color) +
  theme_minimal() + 
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Figure 2: Bond Type Percentage by Most Serious Charge",
    caption = "Central Bond Court Observations, January 2022 to May 2022",
    fill = "Bond Type") +
  xlab("Class of the Most Serious Charge") + 
  ylab("")
```

> "In low-level cases I observed, the pretrial condition was release without money bail in just under half of cases."_ (Page 19)

For class 4 cases, defendants are released on an I-Bond (meaning without money bail) 51% of the time. It is important to note, the results below do not account for cases listed as D-bonds where I did not record the amount the defendant can post or I did not record the amount of money bail set. In `r d_bond_unfiltered - d_bond_filtered` cases bond affordability is undefined and these results do not appear in the final data. This is why, overall, D bond is under-represented in this result and I-Bond is over-represented. In the unfiltered data, I-Bond is slightly less than 50% of cases.

```{r data note in low-level cases}
### PRINTING PERCENTAGES BY CLASS AND BOND TYPE
bond_summary_full %>%
  mutate(percentage = percentage * 100) %>%
  select(class_1, bond_type_detail, percentage) %>%
  pivot_wider(names_from = class_1, values_from = percentage) %>%
  kable(caption = "Table 1: Bond Type Percentage by Most Serious Charge",
        col.names = c("","4", "3", "2", "1", "X"),
        digits = 1)
```

# Gun possession with previous conviction(s)
>_"For example, release on money bail and some type of supervision is the common pretrial condition for defendants charged with gun possession who have a previous gun conviction in their history"_ (Page 20-21)

```{r conditions for gun case with prior record}
### CALCULATING EXACT PERCENTAGE
charge_type_6_total <- obs_data %>%
  filter(charge_type == "6 - gun possession (prior record)"
         & bond_type != "") %>%
  count()

charge_type_6_condition <- obs_data %>%
  filter(charge_type == "6 - gun possession (prior record)"
         & bond_type != ""
         & condition != "") %>%
  count()

### Visualizing results
# using observation data
obs_data %>%
  
  # selecting gun cases where the defendant has a felony record
  filter(charge_type == "6 - gun possession (prior record)"
         & bond_type != "") %>%

  # visualizing results
  ggplot(mapping = aes(
    x = bond_type,
    fill = condition
  )) +
  
  # controlling graph aesthetics
  geom_bar(position = position_dodge2(width = 0.9, preserve = "single")) + 
  scale_fill_manual(values = c("grey85","grey60","grey25")) +
  theme_minimal() + 
  labs(
    title = "Figure 3: Pretrial Conditions for Gun Case with Prior Felony Record",
    caption = "Central Bond Court Observations, January 2022 to May 2022",
    fill = "Pretrial Condition") +
  xlab("Monetary Pretrial Conditions") + 
  ylab("Count of Observations")
```

In Figure 3 we can see pretrial conditions are common in gun cases where the defendant has a prior record. Electronic monitoring and pretrial supervision, combined, are applied in `r charge_type_6_condition / charge_type_6_total * 100`% of cases with a gun charge when the defendant has a prior record. These conditions often overlap with high monetary bail or pretrial jailing as we can see in Figure 3.

> _Of the class X gun possession cases I observed, prosecutors requested pretrial jailing half of the time._ (Page 24-25)

In gun possession cases I observed, pretrial jailing was set in half (7) after prosecutors requested jailing and judges approved the request. In half (7) prosecutors did not request jailing. In every case I observed where prosecutors requested jailing the request was approved.

The "NA" observation represents a case where I observed a class X gun possession case but did not record the pretrial condition set.

```{r class X gun possession pretrial conditions}
# using observation data
obs_data %>%
  
  # selecting class X gun possession case
  # called "armed habitual criminal" which I abbreviated as "AHC" in the charge variable
  filter(charge_1 == "AHC") %>%
  group_by(bond_type) %>%
  summarize(Total = n()) %>%
  
  # printing results
  kable(caption = "Table 2: Count of Bond Type for Class X Gun Possession Cases")
```

# Gun license and release conditions in low-level gun possession cases
> _In a third of class four gun arrests, the defendant is a legal gun owner with a FOID card Illinois gun license_ (Page 28)

In 32% of class 4 gun possession cases, the person charged had a FOID card enabling them to legally own a weapon as we can see in Table 3. The "CCL" Concealed Carry License is needed in addition to a FOID card for some methods of weapon transportation including having an uncased or loaded weapon in a car [FOID law](https://www.ilga.gov/legislation/ilcs/ilcs3.asp?ActID=1657&ChapterID=39) and [CCL law](https://www.ilga.gov/legislation/ilcs/ilcs3.asp?ActID=3497&ChapterID=39).

```{r gun license}
# using observation data
obs_data %>%
  
  # selecting low-level (class 4) gun possession with gun license data recorded
  filter(class_1 == 4, charge_1 == "UUW", gun_license != "") %>%
  
  # counting by license
  group_by(gun_license) %>%
  summarize(Total = n()) %>%
  kable(caption = "Table 3: Gun License Status in Class 4 Gun Possession Cases")
```

> _In the cases I observed, nine of ten defendants with a FOID card are released on an I-Bond, the most lenient pretrial condition. In contrast, only two of ten defendants without a FOID card and arrested on gun charges were released on an I-Bond with most given money bail as a pretrial condition._ (Page 28)

Figure 5 demonstrates the high rate of I-Bonds for defendants with a FOID card and the low rate of I-bonds for defendants without a FOID card. In the count in Table 4 we can see that 3 in 24 defendants with a FOID Card (or 12%) are required to pay money bail while 42 of 50 defendants without a FOID card (or 84%) are required to pay money bail.  

```{r}
# VISUALIZING DATA
# using observation data
obs_data %>%
  
  # selecting low-level gun possession cases
  # narrowing to with or without FOID
  # narrowing to bond information recorded
  filter(class_1 == 4, 
         charge_1 == "UUW", 
         gun_license %in% c("FOID no CCL", "No FOID or CCL"), 
         bond_type != "") %>%
  
  # graphing data
  ggplot(mapping = aes(
    x = gun_license,
    fill = bond_type
  )) +
  
  # controlling graph aesthetics
  geom_bar(position = position_dodge2(width = 0.9, preserve = "single")) +
  scale_fill_manual(values = c("deepskyblue", "grey25")) +
  theme_minimal() + 
  labs(
    title = "Figure 5: Gun License and Pretrial Conditions in Class 4 Gun Possession Cases",
    caption = "Central Bond Court Observations, January 2022 to May 2022",
    fill = "Bond Type") +
  xlab("Gun License") + 
  ylab("Count of Observations")

# COUNTING CASES
obs_data %>%
  
  # selecting low-level gun possession cases
  # narrowing to with or without FOID
  # narrowing to bond information recorded
  filter(class_1 == 4, 
         charge_1 == "UUW", 
         gun_license %in% c("FOID no CCL", "No FOID or CCL"), 
         bond_type != "") %>%
  group_by(gun_license, bond_type) %>%
  summarize(Total = n()) %>%
  kable(caption = "Table 4: Gun License and Pretrial Conditions in Class 4 Gun Possession Cases")
```

# Drug charges pretrial conditions set overview
> _Judges release two-thirds of defendants with drug charges...I observed consistent money bail and pretrial supervision conditions between both charge categories_ (Page 32)

As we see in Figure 6, there are nearly identiical rates of I-bond and D-Bond between drug cases charged as possession and drug cases charged as manufacture and deliver. In both case types, around 65% of defendants are released without money bail and an additional 20% of defendants are released on affordable money bails.

```{r}
### CALCULATING PERCENTAGE OF CASES
# counting by case type and bond type
drug_case_summary <- obs_data %>%
  filter(charge_type %in% c("3 - drug possession (PCS)", 
                            "4 - drug sales (manufacture/delivery, PCS with intent)")
         
         # I did not record bond type for 27 drug case observations 
         & bond_type_detail != "")  %>%
  group_by(charge_type, bond_type_detail) %>%
  summarize(total = n())

# counting by case type
drug_case_type <- aggregate(drug_case_summary$total,
            by = list(charge_type = drug_case_summary$charge_type),
            FUN = sum)

# calculating percentage of bond type grouped by class
drug_case_full <- drug_case_summary %>% 
  left_join(drug_case_type) %>%
  mutate(percentage = total / x)

### VISUALIZING RESULTS
drug_case_full %>%
  ggplot(mapping = aes(
    x = charge_type,
    y = percentage,
    fill = bond_type_detail
  )) + 
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) + 
  scale_fill_manual(values = c("deepskyblue", "grey60", "grey25")) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Figure 6: Money Bail in Drug Cases",
    caption = "Central Bond Court Observations, January 2022 to May 2022",
    fill = "Bond Type") +
  xlab("Case Type") + 
  ylab("Percentatge of Cases")
```

