---
title: "Gun cases per day"
author: "Micah Clark Moody"
date: "3/2/2022"
output: html_document
---

# Load required packages
```{r setup}
library(tidyverse)
library(lubridate)
library(knitr)
```

# Import and clean data
? make import reproducible (get rid of desktop path) (choice based on identifiable data...)
X in .csv remove ASA, atty, and columns after "notes"
- ^ CHANGE, use original .csv and deselect
- vector charge_type to be in order
```{r, setup} 

# importing data from my observations (obs_data)
obs_data <- read.csv("~/Desktop/r/bond_court/data/data_cleaned.csv") %>%
  
  # removing identifiable information
  select(-c(judge))

# coercing numeric variables
obs_data$can_post <- as.numeric(obs_data$can_post) 
obs_data$bond_amount <- as.numeric(obs_data$bond_amount)

# creating dummy variable for excessive bond
bond_excess <- data %>%
  select(date, case, judge, can_post, bond_amount, bond_type, class_1, charge_1, violations, charge_type) %>%
  filter(bond_type == "D") %>%
  drop_na() %>%
  mutate("bond_excess" = ifelse(((can_post * 10) < bond_amount), 1, 0))

#data <- full_join(data, bond_excess)

# CCSAO DATA

initiation <- read.csv("~/Desktop/r/bond_court/data/2022-07-03 CCSAO Data/Initiation.csv")

initiation_test <- initiation %>%
  separate(RECEIVED_DATE, c("RECEIVED_DATE", "time"), " ") %>%
  select(-"time")

test$date <- initiation_test %>%
  mdy("RECEIVED_DATE")

initiation_test$RECEIVED_DATE <- as_date(initiation_test$RECEIVED_DATE) 
  
data$date <- as_date(data$date)

?separate



intake <- read.csv("~/Desktop/r/bond_court/data/2022-07-03 CCSAO Data/Intake.csv")

```

# ASA Data Drug Cases
```{r}
initiation_test %>%
  filter(OFFENSE_CATEGORY == "narcotics")
  
  
# BOND_REDUCED, 1 = yes, 0 = no
initiation_test %>%
  filter(BOND_TYPE_INITIAL == "D") %>%
  
  # not quite right doesn't account for D to I or D to no
  mutate("BOND_REDUCED" = ifelse((BOND_AMOUNT_INITIAL > BOND_AMOUNT_CURRET), 1, 0))

```

Drug cases I observed
```{r}
data %>%
  filter(
    charge_type == "4 - drug sales (manufacture/delivery, PCS with intent)" 
    | charge_type == "3 - drug possession (PCS)") %>%
  group_by(nole) %>%
  summarize(n()) %>%
  mutate("percent" = ((`n()`/(92+16))*100)) %>%
  kable()

data %>%
  filter(
    charge_type == "4 - drug sales (manufacture/delivery, PCS with intent)" 
    | charge_type == "3 - drug possession (PCS)") %>%
  group_by(bond_type) %>%
  summarize(n()) %>%
  mutate("percent" = (`n()`/(32+56)*100)) %>%
  kable()

bond_excess %>%
  filter(
    charge_type == "4 - drug sales (manufacture/delivery, PCS with intent)" 
    | charge_type == "3 - drug possession (PCS)") %>%
  group_by(bond_excess) %>%
  summarize(n()) %>%
  mutate("percent" = (`n()`/(25)*100)) %>%
  kable()


```

# Affordable bond
Bail as a release condition
Intro
In the cases I observed, bail was set at an affordable level % of the time, with a median affordable bail of #
```{r}

# Cases where a D bond was set
data %>%
  group_by(bond_type) %>%
  summarise(n()) %>%
  mutate("percent" = ((`n()`/(439-59))*100)) %>%
  kable()

# Affordability of D bond
bond_excess %>%
  group_by(bond_excess) %>%
  summarise(n()) %>%
  mutate("percent" = ((`n()`/(67+105))*100)) %>%
  kable()

# in cases with violations
bond_excess %>%
  filter(violations == "VOBB" | violations == "IDOC Hold" | violations == "VOP" ) %>%
  group_by(bond_excess) %>%
  summarise(n()) %>%
  mutate("percent" = ((`n()`/(32))*100)) %>%
  kable()

# Median D-Bond
test <- bond_excess %>%
  filter(bond_excess == 0)

median(test$bond_amount)

# excessive bond proportion
mutate(bond_excess$req_post)


post <- bond_excess %>%
  select(bond_amount, can_post) %>%
  mutate(
    req_post = bond_amount / 10
  )


```

Bail as a release condition
4 - AUUW
"This presumption is standard for first time gun possession cases, where bail was set at an affordable level in % of the cases I observed" (07-02-2022)
```{r}

# 4-AUUW cases bond affordable
# NOTE: bond_excess only has 172 obs meaning `235-172` cases where a D-bond was set I didn't record can_post and/or bond_amount
# NOTE: bond_excess data set should be `data` after cleaning the intro section

bond_excess %>%
  filter(class_1 == "4", charge_1 == "UUW") %>%
  group_by(bond_excess) %>%
  summarise(n()) %>%
  mutate("percent" = (`n()`/93)*100) %>%
  kable()

# 4-AUUW cases I-bond

data %>%
  filter(class_1 == "4", charge_1 == "UUW") %>%
  group_by(bond_type) %>%
  summarise(n()) %>%
  mutate("percent" = (`n()`/93)*100) %>%
  kable()
  
```

Bail for jailing
Other violations
```{r}
data %>%
  filter(bond_amount >= 100000) %>%
  group_by(violations) %>%
  summarise(n()) %>%
  mutate("percent" = (`n()`/(18+3+2+11+1))*100) %>%
  kable()

data %>%
  #filter(bond_amount >= 100000) %>%
  group_by(violations) %>%
  summarise(n()) %>%
  mutate("percent" = (`n()`/(297+21+47+21+1+43+9))*100) %>%
  kable()
```

# Drug cases over time
```{r}
offense_category <- unique(intake$OFFENSE_CATEGORY)

narc_initiation <- initiation %>%
  filter(OFFENSE_CATEGORY == "Narcotics")

narc_initiation %>%
  group_by(RECEIVED_DATE) %>%
  summarize(n()) %>%
  ggplot(mapping = aes(
    x = RECEIVED_DATE,
    y = `n()`
  )) +
    geom_point()

```

# Bond initial and current
```{r}

```

# Breaking down data by judge
## Bond type for each judge
Question: What is the percentage use of each bond type by judge?

```{r, bond_type_by_judge}
judge_bond <- data %>%
  filter(judge != "?", bond_type != "") %>%
  group_by(bond_type, judge) %>%
  summarize(n())

judge <- data %>%
  filter(judge != "?", bond_type != "") %>%
  group_by(judge) %>%
  summarise(n()) %>%
  rename("total_cases" = "n()")

judge_bond <- full_join(judge_bond, judge) %>%
  mutate("percent" = `n()` / `total_cases`)

ggplot(judge_bond, mapping = aes(
  x = bond_type,
  y = `percent`
)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(facets = "judge") + 
  labs(
    title = "Percentage of Bond Type by Judge",
    x = "Bond Type",
    y = "Percentage of Total Bonds"
  )

kable(judge)
```

## Excessive bond for each judge
Question: What percentage of cases where judges assign a D-bond is that bond in excess of what defendants can post?

```{r, excessive_bond_by_judge}
bond_excess %>%
  filter(judge != "?", !is.na(bond_excess)) %>%
  ggplot(mapping = aes(
    x = judge,
    y = bond_excess
)) +
  geom_bar(stat = "summary")
```

Bond reductions by case
```{r}
test <- initiation %>%
  mutate()
  
```

# Breaking down EM information
***Limitation:*** Judges do not consistently state or explain EM conditions. For example, "5000 D with a monitor" would be a common ruling but it is not clear the conditions of monitoring. For this reason, I don't distinguish between types of EM. I am also concerned there is an under count of cases with EM so I will not compare EM to the total number of cases, instead limiting my analysis to the cases that are positively identified as having judge-assigned EM.

## Cases with EM information
Question: what cases have EM as a condition? Were these charges legally eligible for pretrial incarceration?

```{r, EM}
# charges that are ELIGIBLE for pretrial incarceration
# importing .csv summarizing probation-eligible charges. See prob_eligible$link for citations
prob_eligible <- read.csv("~/Desktop/r/bond_court/prob_eligible.csv")

# number of EM conditions by charge category
data %>%
  filter(condition == "EM") %>%
  group_by(charge_type) %>%
  ggplot(mapping = aes(
    x = charge_type
  )) +
  geom_bar() +
  scale_x_discrete(guide = guide_axis(angle = 45))

# table of charges
EM <- data %>%
  filter(condition == "EM", class_1 != "Unknown") %>%
  group_by(class_1, charge_1) %>%
  select(class_1, charge_1) %>%
  summarise(total = n()) %>%
  full_join(prob_eligible) %>%
  select(-c("law","link")) %>%
  drop_na()

kable(EM) 

# total incarceration elegible versus not eligible
EM %>%
  ggplot(mapping = aes(
    x = probation_eligible,
    y = total
  )) +
  geom_col()

```

# VOBB cases and charge severity
```{r}
data_cleaned

VOBB <- data_cleaned %>%
  filter(violations == "VOBB") %>%
  group_by(charge_1)

4/38
```
