---
title: "Electronic monitoring pretrial cannot be compared to incarceration"
author: "Micah Clark Moody"
date: '2022-05-04'
output: html_document
---

I've been pondering Gabriela Kirk's insight that prison is so normalized as the response to law violation we think about pretrial electronic monitoring in contrast to incarceration rather than release without conditions [(Kirk 2020)](https://academic.oup.com/socpro/article/68/3/642/5835372?login=true). She writes:

> Individuals’ expectations about their legal options and the power of the state to confine both pre- and post-conviction shaped their discussion of its benefits and detriments. Those who believed jail to be the routine and only alternative to their situation minimized the negative collateral consequences they experienced. However, nearly all respondents experienced significant disruption to their lives, particularly to their employment, housing, and family life...To be sure, there are real benefits to avoiding traditional incarceration. However, ***inventing new modes of confinement outside the walls of jails and prisons does not move away from mass incarceration nor does it eliminate the need to be fully attentive to possible negative collateral consequences*** (Kirk 2021, page 643)

Kirk found "For many on [electronic monitoring], this sanction is, in fact, much more restrictive than
other possible sanctions, such as being out on bond, paying a fine, or attending weekly meetings with
a probation officer" (2021, page 647).

Advocating against electronic monitoring, the Chicago Appleseed Center for Fair Courts has demonstrated electronic monitoring is used as a punishing version of pretrial rather than a altered pretrial incarceration in their [electronic monitoring fact sheet](https://www.chicagoappleseed.org/wp-content/uploads/2021/11/202109_10-Facts-EM-Cook-County-EM-FINAL-updated.pdf).

***My courtwatching aligns with previous findings that comparing pretrial electronic monitoring with pretrial incarceration is a false comparison*** ([Beckett and Murakawa 2012](); [Kilgore 2013](); [Kirk 2020](); [Chicago Appleseed Center for Fair Courts 2021]()). My conclusion is based on 383 cases in Cook County Bond Court on 17 different days. There several limitations to this data, see this repository [README.md](https://github.com/MicahCM/bond_court/blob/main/README.md) for details. The most important limitation to electronic monitoring analysis it that I likely under counted electronic monitoring conditions because I was only verbally listening to proceedings. Judges do not always verbally state when electronic monitoring is a condition of bond and sometimes I was confused and may have missed conditions. This is why I will not compare electronic monitoring rates to total case numbers.

```{r setup,  message = FALSE, warning = FALSE, include = FALSE}
# importing required packages
library(tidyverse)
library(knitr)
```

```{r, data_import, include = FALSE}
# import all courtwatching data
data <- read.csv("~/Desktop/r/bond_court/data_cleaned.csv")

# import probation-eligible cases summary
prob_eligible <- read.csv("~/Desktop/r/bond_court/prob_eligible.csv")

EM <- data %>%
  filter(condition == "EM", class_1 != "Unknown") %>%
  group_by(class_1, charge_1) %>%
  select(class_1, charge_1) %>%
  summarise(total = n()) %>%
  full_join(prob_eligible) %>%
  drop_na()

# changing dummy variable to character
EM$probation_eligible <- ifelse(EM$probation_eligible == 1, "Yes", "No")
```

# Charges involved

I recorded judges including electronic monitoring as a condition of bond in 21 cases. For defendants who were released with electronic monitoring, this table communicates the most serious charge and the total number of cases with that charge.

Naming notes:

- "UUW" refers to possessing a weapon without a required license (in the statue "unlawful use of a weapon")
- "AHC" requires to possessing a weapon with multiple previous convictions (in the statute "armed habitual criminal")
- "PCS" refers to drug possession (in the statute as "possession of a controlled substance")
- "Burglary" was non-residential

```{r, charge_summary, echo = FALSE}
EM %>%
  select(-"probation_eligible") %>%
  drop_na() %>%
  arrange(-total) %>%
  kable(col.names = c("Class", "Charge", "Total Cases"), caption = "Table: Highest charge in cases where electronic monitoring was a bond condition")
```

# The majority of people could not be legally incarcerated

According to Illinois law people can ***only be incarcerated*** pretrial when they are charged with a non-probationable offense listed in [730 ILCS 5/5-5-3](https://www.ilga.gov/legislation/ilcs/fulltext.asp?DocName=073000050K5-5-3). (It is important to note this law will change when a _Pretrial Fairness Act_ provision takes effect on January 1, 2023). I coded each charge as probationable and non-probationable and incarceration was only a definite legal possibility in 5 of 21 cases. 

It is important to note, for some class 2 and 3 charges, it depends on the persons record which I did not record in the data. For example, it is possible a case where the charge is a Class 2 UUW is non-probationable if the defendant has another class 2 conviction from the past 10 years. I choose to code record and enhancement-dependent charges as probationable because the underlying law violation is deemed probationable by the legislature.

```{r, inc_vulnerable, echo = FALSE}
EM %>%
  group_by(probation_eligible) %>%
  summarize(n()) %>%
  ggplot(mapping = aes(
    x = probation_eligible,
    y = `n()`
  )) +
  geom_col(fill = "darkgreen") +
  
  # controlling graph aesthetics
  labs(title = "Is the Highest Charge Non-Probationable?",
    x = "",
    y = "Total cases observed",
    caption = "Source: Micah's courtwatching, see data_cleaned.csv in this repository") +
  theme_minimal()
```

This means pretrial incarceration was not a legal possibility for the majority of people judges assigned to electronic monitoring as apart of their pretrial conditions. 

Further, pretrial incarceration was not an option in any of the cases because incarceration (in almost all cases besides murder charges) requires prosecutors to file a "no bail petition." No bail petitions were not filed in ***a single case*** I have observed where electronic monitoring is a condition of release. Of the 18 no bail petitions I observed, all 18 were granted. This means electronic monitoring was not used as a less restrictive condition or a "prefferred alternative" to incarceration. 

In Illinois today, comparing electronic monitoring to incarceration pretrial is a false comparison. Kirk, Appleseed, and my data all support this finding. We must compare bond with electronic monitoring to release without electronic monitoring. The majority of cases where people are required to submit to electronic monitoring are cases when people cannot be incarcerated and, even in cases where incarceration is legally possible, electronic monitoring is not being used as a less restrictive condition.

### Citations
Beckett, Katherine, and Naomi Murakawa. 2012. “Mapping the Shadow Carceral State: Toward an Institutionally Capacious Approach to Punishment.” _Theoretical Criminology_ 16(2):221–44.

Appleseed.

Kilgore, James. 2013. “Progress or More of the Same? Electronic Monitoring and Parole in the Age of Mass Incarceration.” _Critical Criminology_ 21(1):123–39.

Kirk, Gabriela. 2021. “The Limits of Expectations and the Minimization of Collateral Consequences: The Experience of Electronic Home Monitoring.” _Social Problems_ 68: 642–57.
