---
title: "2017 Analysis: Policy Change in Cook County Bond Court"
author: "Micah Clark Moody"
date: '2022-05-11'
output: github_document
---

This project is to better understand the past impact of bail policy. I'm deeply indebted to the the _Coalition to End Money Bond_ and _Chicago Appleseed Foundation for Fair Courts_ for publishing their cleaned data and code with the [Chicago Data Collaborative](https://gitlab.com/ChicagoDataCooperative) for te public to access. This analysis primarily draws on their insights and is only answering lingering questions I have after largely coming to understand Order 18.8A and its impact through their research, analysis, and reporting. [This report](http://www.chicagoappleseed.org/wp-content/uploads/2018/02/Courtwatching-Report_Coalition-to-End-Money-Bond_FINAL_2-25-18.pdf) is the best starting point for understanding the impact of Order 18.8A in bond court.

***Reader note:*** This a preliminary analysis primarily for my own understanding. If you're hear to learn more about Order 18.8A, I strongly suggest you read the [report](http://www.chicagoappleseed.org/wp-content/uploads/2018/02/Courtwatching-Report_Coalition-to-End-Money-Bond_FINAL_2-25-18.pdf) instead. This will be easier to digest, more detailed, and less likely to have errors. If you're interested in my analysis or the work that went into my MA thesis on bond court, you're in the right place. 

# 2017 Bond Court Policy Change
On September 18, 2017, [General Order 18.8A](https://www.cookcountycourt.org/Manage/Division-Orders/View-Division-Order/ArticleId/2562/GENERAL-ORDER-NO-18-8A-Procedures-for-Bail-Hearings-and-Pretrial-Release) was issued by the Chief Judge changing the rules of bond court to "ensure no defendant is held in custody prior to trial solely because the defendant cannot afford to post bail." This policy change was in response to years of organizing and advocating for an decrease reliance on cash bail. This created a disruption in bond court, requiring facts be presented about a defendants ability to pay and be considered by the judge on the record. Further, for a judge sets bail in excess of what a defendant can pay after September 18, 2017, judges have to put additional findings on the record explaining their decision.

# This data

- [Courtwatching data](https://gitlab.com/ChicagoDataCooperative/court-observation/-/tree/master) collected by volunteer court watchers who physically attended court and recorded observations for each case. 
- [Data description](https://gitlab.com/ChicagoDataCooperative/court-observation/-/blob/master/data_processing.Rmd), and I will also explain the variables relevant to this analysis throughout
- [Bond court summary](https://gitlab.com/ChicagoDataCooperative/bond-court-dispositions) repository created through freedom of information act responses by the Cook County sheriff's office

```{r setup, message = FALSE, warning = FALSE, echo = FALSE}
# importing required packages
library(lubridate)
library(tidyverse)
library(readr)
library(knitr)
library(descr)
```

### Importing data
This data is linked to the GitHub repository for the initial data collection project. The codebooks, .csv files, and data cleaning scripts can be viewed at these links:      
      
- [daily_data](https://gitlab.com/ChicagoDataCooperative/bond-court-dispositions) is the aggregate data from the Cook County sheriff      
- [case_data](https://gitlab.com/ChicagoDataCooperative/court-observation/-/tree/master) is case-level data recorded by court watchers      
        
First, I imported the data, re-code variables as needed, and re-name variables used for analysis using snake case formatting. When the same variable (for example "date" and "bond_type") appear in both data sets they are named the same way to limit confusion about what the variable represents and simplify joining operations.

```{r, message = FALSE, message = FALSE, echo = FALSE}
# tidying daily_data
# importing daily_data
daily_data <- read_csv("https://gitlab.com/ChicagoDataCooperative/bond-court-dispositions/-/raw/master/data/cbc_disposition_stats.csv") %>%
  rename("No_Bond" = "No Bail") %>%
  
# creating a dummy variable for whether order 18.8A was in effect during an observation
  mutate("order_18.8A" = if_else(`date` < "2017-09-18", "before", "after")) %>%
  mutate("order_18.8A" = factor(order_18.8A, levels = c("before", "after")))

# re-coding numeric variable imported as character due to a typo in the data
daily_data$`Not Admitted to EM Prg` <- as.numeric(daily_data$`Not Admitted to EM Prg`)

# tidying case_data
# importing case_data
case_data <- read_csv("https://gitlab.com/ChicagoDataCooperative/court-observation/-/raw/master/aug_sept_oct_by_id.csv") %>%
  rename("date" = "Date", 
         "bond_type" = "Bond_Type",
         "max_charge" = "Max_Charge")

# re-coding "date" column from character to date
case_data_date <- str_split(case_data$date, ", ", n = 2, simplify = TRUE) %>%
  as.data.frame() %>%
  select(-V1)

case_data$date <- mdy(case_data_date$V2)

# creating a dummy variable for whether order 18.8A was in effect during an observation
case_data <- case_data %>%
  mutate("order_18.8A" = if_else(`date` < "2017-09-18", "before", "after")) %>%
  mutate("order_18.8A" = factor(order_18.8A, levels = c("before", "after")))

```

# Distribution of bond types before and after implementation

### Daily bond types
__Question:__ what is the distribution of bond types each day?

__Interpretation:__ there is a clear shift in September when the I-bonds (blue) become more frequent and D-bonds (grey) become less frequent. This suggests Order 18.8A had an effect in decreasing money bail.

The rate of denying bail (red) seems to have increased. The use of I-bonds with electronic monitoring (light blue) appears to have decreased, which is puzzling. The report and people I spoke with suggest this is because new, reform-oriented judges were assigned to bond court.

__Data note:__ I am dropping C-bonds from the analysis. C bonds are incredibly rare, appearing only 61 times in 3045 cases. From my observation of bond court 5 times a week, this matches my expectations, as it is exceedingly rare for a C-bond to be ordered, I observed 1 in over 400 cases of observation. Thus, I will remove C-bonds from my analysis.

```{r bond_type_daily, message = FALSE, echo = FALSE}
# calculating distribution of bond types each day
daily_data <- daily_data %>%

# calculating daily percentages by type
  mutate(
    percent_N = (`No_Bond` / `TOTAL CASES`), 
    percent_I = (`I-Bonds` / `TOTAL CASES`),
    percent_IEM = (`EMI` / `TOTAL CASES`),
    percent_D = (`D-Bonds` / `TOTAL CASES`))

# assigning to one column
bond_type <- daily_data %>%
  pivot_longer(
    cols = starts_with("per"),
    names_to = c(".value", "bond_type"),
    names_sep = 8) %>%
  rename("percent" = "percent_") %>%
  mutate(bond_type = factor(bond_type, levels = c("I", "IEM", "D", "N")))

# creating data visualization
ggplot(data = bond_type, mapping = aes(
    x = date,
    y = percent,
    color = bond_type)) +
  geom_point() +
  geom_smooth() +

# controlling graph aesthetics
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(name = "Bond Type", values = c("cornflowerblue", 
                                                    "lightblue2", 
                                                    "darkgrey", 
                                                    "red")) +
  labs(title = "Figure 1: Daily Cook County Felony Bond Type Distribution",
    x = "Date",
    y = "Percent of total bonds",
    caption = "Source: Chicago Data Collaborative") +
  theme_minimal()
```

### Overall bond types before and after order 18.8A

__Question:__ are D-bonds more prevelant before Order 18.8A is implemented?

__Interpretation:__ yes, there is a significant increase in the percentage of I-Bonds after the order is implemented. Interestingly, there is also an increase in no bonds, jailing pretrial defendants. In future analysis, I am curious if this change is mirrored across charges or if some charges (like drug possession) account for the increase in I-Bonds while others (like more serious charges) account for the increase in No_Bonds.

```{r bond_type_18.8A, echo = FALSE}
# grouping bonds by date before and after 18.8A
bond_type_grouped <- bond_type %>%
  group_by(`order_18.8A`, `bond_type`) %>%
  summarize(mean_percent = mean(percent))

# graphing average percentage
ggplot(data = bond_type_grouped, mapping = aes(
  x = order_18.8A,
  y = mean_percent,
  fill = bond_type)) +
  geom_col(position = "dodge") +

# controlling graph aesthetics
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(name = "Bond Type", values = c("cornflowerblue", 
                                                    "lightblue2", 
                                                    "darkgrey", 
                                                    "red")) +
  labs(title = "Figure 2: Cook County Felony Bond Type Mean",
    x = "Implementation of Order 18.8A",
    y = "Percent of total bonds",
    caption = "Source: Chicago Data Collaborative") +
  theme_minimal()
```

### Descriptive statistics on bond type
From Figure 1 and 2 we see an increase in I-bonds and No bonds after Order 18.8A was implemented. From Figure 1, we also see substantial variation between days. To better appreciate this variation, here are the descriptive statistical characteristics of the data set.

I wanted to figure out descriptive stats before and after the order went into effect. You can see this data visually represented in figure 3.

```{r, echo = FALSE}

print("Quartile summary before Order 18.8A")

# before order quartile summary
daily_data %>%
  filter(order_18.8A == "before") %>%
  select(`percent_N`, `percent_I`, `percent_IEM`, `percent_D`) %>%
  summary()

print("Standard deviation before Order 18.8A")

# before order sd
daily_data %>%
  filter(order_18.8A == "before") %>%
  select(`percent_N`, `percent_I`, `percent_IEM`, `percent_D`) %>%
  sapply(sd)

print("Quartile summary after Order 18.8A")

# after order quartile summary
daily_data %>%
  filter(order_18.8A == "after") %>%
  select(`percent_N`, `percent_I`, `percent_IEM`, `percent_D`) %>%
  summary()
 
print("Standard deviation after Order 18.8A")
 
# after order sd
daily_data %>%
  filter(order_18.8A == "after") %>%
  select(`percent_N`, `percent_I`, `percent_IEM`, `percent_D`) %>%
  sapply(sd)

# box plot
ggplot(bond_type, mapping = aes(
  x = order_18.8A,
  y = percent,
  fill = bond_type
)) + 
  geom_boxplot() +
  
  # controlling graph aesthetics
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(name = "Bond Type", values = c("cornflowerblue", 
                                                    "lightblue2", 
                                                    "darkgrey", 
                                                    "red")) +
  labs(title = "Figure 3: Cook County Felony Bond Type Distribution",
    x = "Implementation of Order 18.8A",
    y = "Percent of total bonds",
    caption = "Source: Chicago Data Collaborative") +
  theme_minimal()

```

# Proportion of no bond rulings before and after order 18.8A
__Question:__ What cases experienced a increase in the proportion of no bond orders after the amount of cash bail possible was constrainted by order 18.8A?

__Interpretation:__ Here, we can see that changing response to serious crimes drove the increase in no-bail rulings. The starkest increase is in murder charges, with all cases having bond denied after September 18th. Robbery, aggrevated battery, crimes against persons, and gun possession are the other 4 charges to experience a significant increase in No_Bonds and a corresponding decrease in D-Bonds.

```{r bond_proportion_by_charge, message = FALSE, warning = FALSE, echo = FALSE}
# grouping by charge and finding proportion of each bond type
bond_by_charge <- case_data %>%
  
# selecting relevant variables
  select(bond_type, max_charge, order_18.8A) %>%
  drop_na() %>%
  filter(bond_type != "C") %>%
  group_by(max_charge, order_18.8A, bond_type) %>%
  summarise(total = n()) %>%
  
# creating count of each bond type
  pivot_wider(names_from = bond_type,
              values_from = total) %>%
  mutate_all(funs(ifelse(is.na(.), 0, .))) %>%
  mutate(total = sum(`I-bond`, `D-bond`, `No Bond`)) %>%

# calculating percentage of each bond type
  mutate(percent_N = (`No Bond` / total),
    percent_D = (`D-bond` / total),
    percent_I = (`I-bond` / total)) %>%
 pivot_longer(cols = starts_with("per"),
    names_to = c(".value", "bond_type"),
    names_sep = -1) %>%
  rename("percent" = "percent_") %>%
  mutate(bond_type = factor(bond_type, levels = c("I", "D", "N")))

# graphing data
ggplot(bond_by_charge, mapping = aes(
  x = bond_type,
  y = percent,
  fill = order_18.8A)) +
  geom_col(position = "dodge") +
  facet_wrap(vars(max_charge)) +
  
# controlling graph aesthetics
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(name = "Order 18.8A in Effect", values = c("azure3", "azure4")) +
  labs(title = "Figure 4: Bonds Type by Charge",
       subtitle = "Before and after order 18.8A went into effect",
       y = "Percentage of Total Bonds",
       x = "Bond Type",
       caption = "Source: Chicago Data Collaborative") +
  theme_minimal()
```

# Charges with the highest porportion of no bond ruling

__Question:__ what charges have the highest rate of bond being denied?

__Interpretation:__ serious felonies, primarily murder, have the highest rates of bond being completely denied. This makes sense, given the condition to deny bond is a finding the defendant posses a danger to an individual or the community. Murder, specifically, is a mandatory no bond due to [this law]().

```{r, echo = FALSE}
# isolating no bond after 18.8A in effect
bond_by_charge %>%
  filter(bond_type == "N", order_18.8A == "after") %>%
  
# graphing data
  ggplot(mapping = aes(
    x = reorder(max_charge, percent),
    y = percent)) +
  geom_col(fill = "azure4") +
  coord_flip() +
  
# controlling graph aesthetics
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Figure 5: Percent No Bond by Charge",
       subtitle = "After order 18.8A went into effect",
       y = "Percentage of Bonds by Charge",
       x = "Maximum Charge in Case",
       caption = "Source: Chicago Data Collaborative") +
  theme_minimal()
```


# Conclusions

### Takeaways from this data
The implementation of order 18.8A on September 18, 2017, appears to have a significant impact on the bonds granted to defendants charged with felonies in Cook County. In Figure 3, we see that before order 18.8A D-Bonds, where the defendant posts money, are most common while after the order I-Bonds, where the defendant does not post money, are more common.

At the same time, there is an increase of denying bond, where there is no possibility the defendant will be released. This is concentrated in the most serious charges but still poses concerns for freedom and due process for defendants charged with the most serious crimes.

### See the Coalition to End Money Bond's [excellent report](http://www.chicagoappleseed.org/wp-content/uploads/2018/02/Courtwatching-Report_Coalition-to-End-Money-Bond_FINAL_2-25-18.pdf) for more information
